---
title: "Remove all error sequences results"
author: "Hannah Swan"
date: "December 5, 2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
```

```{r}
source("/scratch/hswan/thesis_isomiR_count_denoising/denoise_isomiR_counts_WORKING_FUNCTION.R")
```

```{r}
mousedata = load_mouse_miRNA_data()
rowdata = mousedata$rowdata
countdata=mousedata$countdata
count_df = rowSums(countdata)
transition_probs = readRDS("/scratch/hswan/thesis_isomiR_count_denoising/transition_probs.Rds")
```

```{r}

files = list.files("/scratch/hswan/thesis_isomiR_count_denoising/rmv_all_err_seqs")
files[1]

#get the files for miRNAs that attempted to have all error sequences removed 

#load partition_obj from all these files:
partition_obj_1 = paste0("/scratch/hswan/thesis_isomiR_count_denoising/rmv_all_err_seqs/", files[[length(files)]], collapse="") %>% readRDS()

miRNA = partition_obj_1$partition_df$miRNA_name[1]
cat("Partition obj for miRNA,", miRNA, "\n")

cat("Partitioning algorithm ran for", partition_obj_1$niter-1, "iterations \n")

cat("Partitions and the amount of sequences in each one:\n")
table(partition_obj_1$partition_df$partition)

cat("Center sequence of each partition and associated observed read count across all samples \n")
filter(partition_obj_1$partition_df, center == 1)

obs_read_counts_table = data.frame(table(partition_obj_1$partition_df$count)) 
colnames(obs_read_counts_table) = c("count",  "freq")
ggplot(obs_read_counts_table, aes(x=count, y=freq)) + geom_point() + ggtitle("Frequency of obs read count vs. obs read count")
```

## Load all partition objects 
```{r}
file_paths = sapply(files, function(x) paste0("/scratch/hswan/thesis_isomiR_count_denoising/rmv_all_err_seqs/", x)) 
partition_objs = lapply(file_paths, readRDS)
```

\n\n For each partition object, get the number of iterations it ran for 
```{r}
num_iters_ran = vector(length=length(file_paths))

for(i in 1:length(partition_objs)){
  num_iters_ran[i] = partition_objs[[i]]$niter
}

cat("Number of iterations run to remove all error sequences:\n")
table(num_iters_ran)
cat("Mean number of iterations run to remove all error sequences:", mean(num_iters_ran), "\n")
```

\n\n Also for each partition, get the number of partitions created
```{r}
num_partitions = lapply(partition_objs, function(x) return(max(x$partition_df$partition))) %>% unlist()
cat("Number of true isomiR sequences identified for each miRNA \n")
table(num_partitions)
```

\n\n Also curious what the proportion of true isomiR sequences identified by our method to the total number of sequences mapping to that miRNA is: 
```{r}
prop_true_isomiR = vector(length=length(partition_objs))
for(i in 1:length(partition_objs)){
  num = max(partition_objs[[i]]$partition_df$partition)
  denom = nrow(partition_objs[[i]]$partition_df)
  prop_true_isomiR[i] = num/denom
}


head(prop_true_isomiR)
cat("Minimum proportion of true isomiR sequences identified:", min(prop_true_isomiR), "\n")
cat("Maximum proportion of true isomiR sequences identified:", max(prop_true_isomiR), "\n")
```

## Curious to see if functions works when we specify we want to use Bonferroni for multiple testing 
```{r}
unique_miRNAs = unique(rowdata$miRNA_name)
num_seqs_mapped = sapply(unique_miRNAs, function(x) filter(rowdata, miRNA_name ==x) %>% nrow())
miRNA = num_seqs_mapped[num_seqs_mapped >= 10] %>% sort() %>% names()
miRNA = miRNA[1]


cat("Testing function with Bonferroni adjustment for miRNA", miRNA, "\n")

tst = denoise_isomiR_counts(rowdata, count_df, transition_probs, miRNA, 0.05, 5, "Bonferroni")
tst_BH = denoise_isomiR_counts(rowdata, count_df, transition_probs, miRNA, 0.05, 5, "BH")
```

\n\n Compare the 2
```{r}
cat("Bonferroni:\n")
table(tst$partition_df$partition)
cat("Benjamini-Hochberg:\n")
table(tst_BH$partition_df$partition)
```
\n\n So same number of partitions results from using both methods for adjusting for multiple testing. 

\n\n Number of iterations run for both:
```{r}
cat("Algorithm with Bonferroni adjustment for multiple testing ran for:", tst$niter-1, "iterations\n")
cat("Algorithm with Benjamini-Hochberg adjustment for multiple testing ran for:", tst_BH$niter-1, "iterations \n")

cat("Do we end up with the exact same partitions from the two methods?\n")
if(all.equal(tst$partition_df$partition, tst_BH$partition_df$partition) == TRUE){
  cat("Yes\n")
}
```

\n \n Small number of sequences mapping to this miRNA so we can look at all rows in the partition_df object
```{r}
tst$partition_df

```

Look at the sequences and the alignments in each partition - are length variants / mixed type isomiRs more likely to get their own partition while sequence variants are more likely to stay in the same partition? 

```{r, echo = TRUE}
#look at partition 1 - 
part_1_subset = filter(tst$partition_df, partition == 1) 

Biostrings::pairwiseAlignment(pattern = part_1_subset$uniqueSequence[part_1_subset$center==1], part_1_subset$uniqueSequence[1])
Biostrings::pairwiseAlignment(pattern = part_1_subset$uniqueSequence[part_1_subset$center==1], part_1_subset$uniqueSequence[2])
Biostrings::pairwiseAlignment(pattern = part_1_subset$uniqueSequence[part_1_subset$center==1], part_1_subset$uniqueSequence[3])

lambda = tst$master_lambdas[[1]][[part_1_subset$uniqueSequence[3]]]
nj = part_1_subset$count[part_1_subset$center == 1]
ni = part_1_subset$count[part_1_subset$uniqueSequence== part_1_subset$uniqueSequence[3]]

#note that the following 2 lines are equivalent and dividing 1 by the other will always give 1. So we will never give a singleton sequence it's own partition under this model 
ppois(0, nj*lambda, lower.tail=F)
1-dpois(0, nj*lambda)

#look at pairwise alignments between initial center sequence and the other center sequences: 
center_seqs = tst$partition_df %>% filter(., center == 1) %>% select(., uniqueSequence) %>% unlist()
initial_center_seq = tst$initial_center_seq
center_seqs = center_seqs[center_seqs != initial_center_seq]

Biostrings::pairwiseAlignment(pattern = initial_center_seq, center_seqs[1])
Biostrings::pairwiseAlignment(pattern = initial_center_seq, center_seqs[2])
Biostrings::pairwiseAlignment(pattern = initial_center_seq, center_seqs[3])
```
