---
title: "Untitled"
author: "Hannah Swan"
date: "December 4, 2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
```

```{r}
source("/scratch/hswan/thesis_isomiR_count_denoising/denoise_isomiR_counts_WORKING_FUNCTION.R")
library(tidyverse)
```

## Load the partition objects that we have
```{r}
filepath = "/scratch/hswan/thesis_isomiR_count_denoising/test_run_isomiR_count_denoising"
files = list.files(filepath)
files = lapply(files, function(x) paste0("/scratch/hswan/thesis_isomiR_count_denoising/test_run_isomiR_count_denoising/", x, collapse=""))
files = sapply(files, unlist)
partition_obj = lapply(files, readRDS)
```

## Look at first partition_obj we loaded to see what we've got  
```{r}
miRNA = strsplit(files[1], "/")[[1]][6] %>% strsplit(., "_test_run_isomiR_count_denoising.Rds") %>% unlist()
cat("partition_df object for", miRNA, "\n")

partition = partition_obj[[1]]$partition_df

cat("Algorithm ran for", partition_obj[[1]]$niter-1, "iterations on miRNA.\n")
#Look at the first few rows:
head(partition)

cat("Number of sequences in each partition:\n")
table(partition$partition)
cat("Algorithm identified", max(partition$partition), "unique true isomiR sequences for miRNA", miRNA, "\n")
cat("There were", nrow(partition), "total sequences mapping to this miRNA across all samples.\n")
```

## Now move to looking at things like the observed read counts of the center sequences
```{r}
filter(partition, center == 1)
s = filter(partition, partition == 5) %>% filter(., center == 0) %>% select(., uniqueSequence) %>% unlist()
filter(partition, partition==5)
c = partition$uniqueSequence[partition$center == 1 & partition$partition == 5] 
lapply(s, Biostrings::pairwiseAlignment, pattern = c)

```

## Look at the center sequences for the first 5 partitions:
```{r}
filter(partition, partition >= 1 & partition <= 5) %>% filter(., center == 1)
cat("Pairwise alignments between center sequences 2-5 and the initial center sequence\n")
s = filter(partition, partition >= 2 & partition <= 5) %>% filter(., center == 1) %>% select(., uniqueSequence) %>% unlist()
c = partition$uniqueSequence[partition$center == 1 & partition$partition ==1]
lapply(s, Biostrings::pairwiseAlignment, pattern = c)
```

## For each partition object, get the number of iterations that ran (i.e. did any of the miRNAs have all their true isomiRs id'd in 5 iterations or less??)
```{r}
num_iters = vector(length=length(partition_obj))
for(i in 1:length(partition_obj)){
  num_iters[i] = partition_obj[[i]]$niter
}

#ok so most of the miRNAs would have needed more iterations to remove all the 
table(num_iters)

num_partitions_created = vector(length=length(partition_obj))
  for(i in 1:length(partition_obj)){
    num_partitions_created[i] = partition_obj[[i]]$partition_df$partition %>% max()
  }
table(num_partitions_created)

num_seqs_mapped = vector(length = length(partition_obj))
for(i in 1:length(partition_obj)){
  num_seqs_mapped[i] = nrow(partition_obj[[i]]$partition_df)
}
table(num_seqs_mapped)
```


## Need to look more closely at the results from miRNA that created partition_obj number 57 - only created one partition but not sure that was right 
```{r}
miRNA = strsplit(files[81], ".Rds") %>% unlist() %>% strsplit(., "/scratch/hswan/thesis_isomiR_count_denoising/test_run_isomiR_count_denoising/")
miRNA = miRNA[[1]][[2]] %>% strsplit(., "_test_run_isomiR_count_denoising") %>% unlist()
transition_probs = readRDS("/scratch/hswan/thesis_isomiR_count_denoising/transition_probs.Rds")

mousedata = load_mouse_miRNA_data()
rowdata = mousedata$rowdata
countdata = mousedata$countdata
count_df=rowSums(countdata)

tst = denoise_isomiR_counts(rowdata, count_df, transition_probs, miRNA, 0.05, 5, "BH")
```

```{r}
#only running for one iteration - pul out results_df:
results_df = tst$results_df
inf_raw_p_value_seqs =  filter(results_df, raw_p == Inf) %>% row.names()
center_seq = tst$partition_df$uniqueSequence[tst$partition_df$center == 1]

Biostrings::pairwiseAlignment(patter=center_seq, inf_raw_p_value_seqs[1])
filter(tst$partition_df, uniqueSequence == inf_raw_p_value_seqs[1])

#insanely small lambda makes sense when we're looking at a weird alignment like this 
lambda = tst$master_lambdas[[1]][[inf_raw_p_value_seqs[1]]]
cat("Lambda:", lambda, "\n")

nj = filter(tst$partition_df, uniqueSequence==center_seq) %>% select(., count) %>% unlist()
cat("nj:", nj, "\n")

#dividing by 0
ppois(0, lambda*nj, lower.tail=F)
dpois(0, lambda*nj)


#why is the function not correctly IDing a new center sequence? 
unique_partitions = unique(tst$partition_df$partition) %>% sort()

#partition isomiRs:
partition_isomiRs = filter(tst$partition_df, partition == 1 & center == 0) %>% select(., uniqueSequence) %>% unlist() %>% unname()
#get significant_seqs
significant_seqs = filter(results_df, sig==1) 
#get new center sequence, may need additional filtering step if there's a tie for sequence with the smallest p-value 
new_center_seq = filter(significant_seqs, adj_p == min(adj_p))  %>% row.names()
new_center_seq = filter(tst$partition_df, uniqueSequence %in% new_center_seq) %>% filter(., count==max(count)) %>% select(., uniqueSequence) %>% unlist()

#create new partition
new_partition = max(tst$partition_df$partition)+1
update_df = tst$partition_df


update_df$partition[update_df$uniqueSequence==new_center_seq] =  new_partition
update_df$center[update_df$uniqueSequence==new_center_seq] = 1

significant_seqs = filter(results_df, sig == 1) %>% row.names()
center_seqs = update_df$uniqueSequence[update_df$center == 1]
significant_seqs=significant_seqs[which(!(significant_seqs %in% center_seqs))]

new_alignments = lapply(significant_seqs, Biostrings::pairwiseAlignment, pattern=new_center_seq)
names(new_alignments) = significant_seqs
new_transitions = lapply(new_alignments, get_transitions)
new_lambdas = lapply(new_transitions, compute_lambda, transition_probs=transition_probs)

master_lambdas = list(tst$master_lambdas[[1]], new_lambdas)
names(master_lambdas) = c(1,2)
for(i in significant_seqs){
  l = lapply(master_lambdas, function(x) return(x[[i]])) %>% unlist()
  new_p = names(which(l==max(l))) %>% as.numeric()
  if(length(new_p) >1){
    new_p = sample(new_p, 1)
  }
  update_df$partition[update_df$uniqueSequence==i] = new_p
}
```